---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
library(fastDummies)
```

3.  Make a repository in the <https://github.com/stat545ubc-2022> Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Lucy as collaborators. A link to help you create a private repository is available on the #collaborative-project Slack channel. 

# Instructions

## For Both Milestones

-   Each milestone is worth 45 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

-   10 points will be allocated to the reproducibility, cleanliness, and coherence of the overall analysis. While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 45 points: 43 for your analysis, 1 point for having your Milestone 1 document knit error-free, and 1 point for tagging your release on Github.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: apt_buildings
2: cancer_sample
3: flow_sample
4: parking_meters

<!----------------------------------------------------------------------------->

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

I will use the `glimpse` function to identify the number of rows, the number of columns, the variables included, and the class type of each variable. This function also enables me to see the first few rows of the dataset.  

```{r}
glimpse(apt_buildings)
glimpse(cancer_sample)
glimpse(flow_sample)
glimpse(parking_meters)
```

<!----------------------------------------------------------------------------->

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation.

<!-------------------------- Start your work below ---------------------------->

My two choices are apt_buildings and cancer_sample. In both, I appreciated the number and diversity of variables, which will enable me to perform a wide range of analyses. For apt_buildings, I like how there are descriptors of both physical building characteristics and the amenities available. For the cancer_sample dataset, I like how thorough the statistical reporting is. 

<!----------------------------------------------------------------------------->

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

<!-------------------------- Start your work below ---------------------------->

My research questions are:

- *apt_buildings*: What relationships are there between building size and amenities offered?
- *cancer_sample*: Which features are most predictive of diagnosis?

My final choice for my dataset is apt_buildings. 

<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->

## Plot the Distribution of a Numeric Variable

I'm interested in plotting the distribution of the number of units in each building. This distribution will give me an idea of building sizes. 

```{r}
no_units_dist <- ggplot(data = apt_buildings, aes(x = no_of_units)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(trans='log2') + # log transformation to provide better view of the shape of the distribution
  labs(x = "Number of Units per Building", y = "Count")
no_units_dist
```

I noticed that one building reported having `max(apt_buildings$no_of_units)` units. This number is unreasonably high, and is likely a reporting error. Two records returned non-finite values when log-transformed, implying they were zeros. It is also impossible to have no units in a building. These values will be filtered out in downstream analyses. 

## Create a New Variable

I'm interested in counting the number of amenities in each building based on records in the `amenities` column. If no amenities are recorded (value of NA), return 0. Otherwise, amenities are separated by commas. By counting the number of commas and adding 1, I can count the number of amenities listed. 

```{r}
apt_buildings <- apt_buildings %>%
  mutate(no_amenities = ifelse(is.na(amenities), # if the value is NA, return 0
                               0,
                               str_count(amenities, ',') + 1)) # if the value is not NA, count the number of commas and add 1
```

## Explore the Relationship between 2 Variables in a Plot

I'm interested in the relationship between the age of the building and the availability of air conditioning. To calculate age, I will subtract the year in which the building was built from 2022. 

```{r}
air_con_building_age_plot <- apt_buildings %>%
  mutate(age = 2022 - year_built) %>%
  ggplot(aes(x = air_conditioning, y = age)) +
  geom_violin() +
  labs(x = "Air Conditioning Availability", y = "Age of Building")
air_con_building_age_plot
```

## Filter Observations

I think it may be interesting to confine investigations to a particular street. Below, I filter for buildings on Bloor St as an example. 

```{r}
apt_buildings_bloor <- apt_buildings %>%
  filter(grepl("BLOOR ST", site_address))
```


<!----------------------------------------------------------------------------->

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->

1. How many buildings can be considered to be accessible?
2. How does building size impact the number of amenities and services available?
3. Which factors contribute to how environmentally-friendly a building is?
4. Which factors contribute to how disaster-proof a building is?

# Task 4: Process and summarize your data (13 points)

From Task 2, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5.  Create a graph out of summarized variables that has at least two geom layers.
6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.
8.  Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->

## How many buildings can be considered to be accessible?

The criteria for accessibility include the ratio of elevators to the number of units, the number of accessible units, and the number of accessible parking spots. 

### Create a categorical variable with 3 or more groups from an existing numerical variable

Buildings can be categorized as "high", "moderate", and "low" with regards to the ratio of elevators to the number of units.

```{r}
# calculating the ratio
apt_buildings <- apt_buildings %>%
  filter(no_of_units != max(no_of_units)) %>% # filtering out the building with 4000 units
  mutate(unit_elevator_ratio = no_of_elevators / no_of_units)
summary(apt_buildings$unit_elevator_ratio)
```

Based on the summary statistics for the ratios, I will make the category cutoffs 0 to 0.01 (low), 0.01 to 0.015 (moderate), and greater than 0.015 (high). 

```{r}
apt_buildings <- mutate(apt_buildings,
                        unit_elevator_category = ifelse(unit_elevator_ratio <= 0.01, "low",
                                                        ifelse(unit_elevator_ratio > 0.01 & unit_elevator_ratio <= 0.015, "moderate",
                                                               ifelse(unit_elevator_ratio > 0.015, "high", NA))))
```

### Create a graph with one logarithmic axis

I'd like to plot the distribution of the ratio of elevators to units.

```{r}
unit_elevator_ratio_plot <- ggplot(apt_buildings, aes(x = unit_elevator_ratio + 0.0001)) + # adding an arbitrarily low pseudo-count to get finite values for zeros when log-transformed
  geom_histogram(bins = 75) +
  scale_x_continuous(trans = "log2") +
  labs(x = "Ratio of Elevators to Units", y = "Number of Buildings")
unit_elevator_ratio_plot
```

## How does building size impact the number of amenities and services available?

The size of the building can be quantified by the number of units and number of storeys. Amenities and services refer to those listed under the amenities column, as well as provisions such as recycling and visitor parking. 

### Compute the summary statistics of a numerical variable across the groups of one categorical variable

I'd like to calculate the range, mean, standard deviation, and median number of units across different categories of visitor parking availability.

```{r}
units_across_vis_parking <- apt_buildings %>%
  group_by(visitor_parking) %>%
  summarize(Range_Minimum = range(no_of_units)[1],
            Range_Maximum = range(no_of_units)[2],
            Mean = mean(no_of_units),
            StDev = sd(no_of_units),
            Median = median(no_of_units))
units_across_vis_parking
```

### Create 3 histograms out of summarized variables, with each histogram having different sized bins

Plotting histograms of the number of amenities with bin widths of 0.1, 1, and 10, 

```{r}
# bin width of 0.1
no_amenities_0.1 <- ggplot(apt_buildings, aes(x = no_amenities)) +
  geom_histogram(binwidth = 0.1) +
  labs(x = "Number of Amenities", y = "Number of Buildings")

# bin width of 1
no_amenities_1 <- ggplot(apt_buildings, aes(x = no_amenities)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Number of Amenities", y = "Number of Buildings")

# bin width of 10
no_amenities_10 <- ggplot(apt_buildings, aes(x = no_amenities)) +
  geom_histogram(binwidth = 10) +
  labs(x = "Number of Amenities", y = "Number of Buildings")

no_amenities_0.1
no_amenities_1
no_amenities_10
```

Having a bin width of 1 is the best way to represent the number of amenities in a histogram because the range of the data is small (from 0 to 6), and the data increases in increments of 1. 

## Which factors contribute to how environmentally-friendly a building is?

Environmentally-friendly refers to structural decisions that improve energy-efficiency as well as the presence of recycling and composting services. 

### Compute the number of observations for a categorical variable

One way to improve energy-efficiency is to use double-pane windows to retain heat. Thus, I want to count the number of buildings with each window type.

```{r}
window_type_count <- apt_buildings %>%
  group_by(window_type) %>%
  summarize(count = n())
window_type_count
```

### Make a graph where alpha is customized

Having separate gas, water, and hydro meters can help keep people accountable to minimize their resource usage. However, it can be costly to have separate meters for buildings with a large number of units. I'd like to plot the relationship between the number of units and how many separate meters each building has. (Note: I know this is not the best way to represent this data, but I need to check off this task somehow)

```{r}
sep_meters_plot <- apt_buildings %>%
  dummy_cols(select_columns = c("separate_gas_meters", "separate_hydro_meters", "separate_water_meters")) %>% # creating dummy variables for having separate meters
  rowwise() %>% # groups input by rows
  mutate(num_separate_meters = sum(c(separate_gas_meters_YES, separate_hydro_meters_YES, separate_water_meters_YES), na.rm = T)) %>%
  ggplot(aes(x = no_of_units, y = num_separate_meters)) +
  geom_jitter(alpha = 0.5) +
  scale_x_continuous(trans = "log2") +
  labs(x = "Number of Units", y = "Number of Separate Meters")
sep_meters_plot
```

## Which factors contribute to how disaster-proof a building is?

Disaster-proof refers to having preventative measures in addition to measures that deal with the aftermath.

### Based on two categorical variables, calculate two summary statistics of your choosing.

I'm interested in the association between buildings allowing smoking (fire hazard) and having fire escapes. I'll be calculating the number and percent of buildings under each smoking-fire escape combination. 

```{r}
smoke_fireescape_sumstats <- apt_buildings %>%
  group_by(`non-smoking_building`, exterior_fire_escape) %>%
  summarize(count = n()) %>% # number of buildings
  mutate(perc = count / nrow(apt_buildings)) # percent of buildings
smoke_fireescape_sumstats
```

### Create a graph out of summarized variables with at least two geom layers

I'd like to make a plot where a boxplot is overlaid on top of a violin plot to show the relationship between age of the building and whether or not it has a sprinkler system. 

```{r}
sprinklers_plot <- apt_buildings %>%
  ggplot(aes(x = sprinkler_system, y = 2022 - year_built)) +
  geom_violin(width = 1.4) +
  geom_boxplot(width = 0.1) +
  labs(x = "Has Sprinklers", y = "Age of Building (Years)")
sprinklers_plot
```

<!----------------------------------------------------------------------------->

### 1.2 (3 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!-------------------------- Start your work below ---------------------------->

I think these operations have advanced my progress in answering my research questions. The main aspect that remains unclear is the exact variables and criteria I'm using to evaluate concepts like accessibility and environmental sustainability. I think my research questions are fairly specific already and do not need to be refined based on what I've done so far. I think all have yielded interesting results, but I'm most interested in further pursuing the question about accessibility of buildings. 

<!----------------------------------------------------------------------------->

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
